---
title: "Shipping Optimization in R"
author: "Connor Birkhold"
date: "4/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
# clear workspace
rm(list = ls())
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(lpSolveAPI)
```

##### Load the data
```{r}
orderData <- fread("./data/order_data.csv")
serviceData <- fread("./data/shipping_service_data.csv")

orderData$order_id <- as.factor(orderData$order_id)
serviceData$delivery_service <- as.factor(serviceData$delivery_service)

```

#### Order Data
```{r}
head(orderData)
str(orderData)
```


#### Service Data
```{r}
head(serviceData)
str(serviceData)
```

#### Add features
* Adding a `days_to_deliver` feature to the orderData to communicate how many days we have to deliver the order
* Adding a `package_cnt` variable to the orderData to identify number of packages. For now each order is assumed to be shipped in one package
* Adding a `total_miles` feature to serviceData to communicate how far a distance the service can cover
```{r}
# order data
orderData <- orderData[,
                       `:=`(date_ordered = as.Date(date_ordered, "%m/%d/%Y"),
                            promised_delivery_date = 
                                as.Date(promised_delivery_date, "%m/%d/%Y"))]

orderData <- orderData[, days_to_deliver := difftime(promised_delivery_date, date_ordered,
                                                     units = 'days')]

orderData$days_to_deliver <- as.numeric(orderData$days_to_deliver)

orderData$package_cnt <- 1

# service data
serviceData <- serviceData[, total_miles := days_in_transit * miles_covered_per_day]
```

####
```{r}


shipping.opt <- function(order_data, service_data, carrier){
    
    # set up a matrix to output results into
    colCnt <- nrow(unique(serviceData[serviceData$delivery_service == carrier,
                                      "service_type"]))
    
    result_matrix <- matrix(0, nrow = nrow(orderData),
                            ncol = colCnt)
    
    row.names(result_matrix) <- orderData$order_id
    
    servTypes <- unique(serviceData[serviceData$delivery_service == carrier, 
                            "service_type"])
    
    colnames(result_matrix) <- c(servTypes$service_type)
    Cost <- matrix(0, nrow = nrow(orderData), ncol = 1)
    colnames(Cost) <- "Cost"
    result_matrix <- cbind(result_matrix, Cost)
        
    # for loop to solve optimization problem
        for(i in 1:nrow(order_data)){
            
            # start with 0 constraints and 3 decision variable
            lprec <- make.lp(nrow = 0, 
                             ncol = nrow(service_data[service_data$delivery_service == carrier,
                                                      "cost_per_package"]))
            
            # select minimize
            lp.control(lprec, sense = "min")
            
            # set type to integer
            set.type(lprec, 
                     columns = 1:nrow(service_data[service_data$delivery_service == carrier, 
                                                   "cost_per_package"]),
                     type = "integer")
            
            # set objective function coefficients
            coef <- service_data[service_data$delivery_service == carrier,
                                 "cost_per_package"]
            set.objfn(lprec, 
                      obj = coef$cost_per_package)
            
            # days_in_transit must be less than or equal to days_to_deliver
            daysToDel <- service_data[service_data$delivery_service == carrier,
                                               "days_in_transit"]
            add.constraint(lprec, 
                           xt = daysToDel$days_in_transit,
                           type = "<=", rhs = c(order_data[i, "days_to_deliver"]))
            
            # must select one service
            selectOne <- rep(1, length(1:length(daysToDel$days_in_transit)))
            
            add.constraint(lprec, xt = selectOne,
                          type = "=", rhs = 1)
            
            # solve the lp
            solve(lprec)
            
            # store results in result_matrix
            matColcnt <- ncol(result_matrix) - 1
            result_matrix[i, 1:matColcnt] <- get.variables(lprec)
            result_matrix[i, "Cost"] <- get.objective(lprec)
        }
    
    return(result_matrix)
    
}
        
fedExCost <- shipping.opt(order_data = orderData,
                          service_data = serviceData,
                          carrier = "fedex")

upsCost <- shipping.opt(order_data = orderData,
                          service_data = serviceData,
                          carrier = "ups")

uspsCost <- shipping.opt(order_data = orderData,
                          service_data = serviceData,
                          carrier = "usps")

```

